---
title: Trial of HiTME package with whole tumor datasets
author: "J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---


# Prepare packages and environtment
```{r load_packages_functions, message=F, warning=F}
renv::restore()

# load packages
pckgs <- c("tidyverse", "Seurat", "scGate", 
           "Matrix", "patchwork",
           "data.table", "STACAS",
           "SignatuR", "ProjecTILs",
           "scCustomize", "UCell")
for(package in pckgs){
  tryCatch(
    {
      library(package,
              character.only = T,
              quietly = T)
    },
    error = function(e){
      message(e)
    })
}

options(timeout=4000) # increase time out for downloading

source("R/utils.R")
source("R/main.R")


my_seed <- 321
set.seed(my_seed)
```

Also, increase memory parameters to allocate large vectors, otherwise R might crash
```{bash create_Renviron}
#include vector max size
max_size=250Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Checking size of the allowed vector:
  cat .Renviron
fi
```




# Load data
We will use the datasets obtained from https://immucanscdb.vital-it.ch/ and previously processed. Only using the ones with whole tumor cells. Moreover, we will also add the Joanito et al., 2022 dataset which include whole tissue from colorectal samples.


The list of datasets from https://immucanscdb.vital-it.ch/ are detailed in https://docs.google.com/spreadsheets/d/1bFZ_jnmWUDJ9gaycsLaIsRmUryUWpdpW5aZY7V4Vpc0/edit#gid=0

We retrieve this dataframe from google drive

```{r}
imm.db <- readxl::read_xlsx("data/Standarized_datasets_DB.xlsx")

# keep whole tumor datasets
table(imm.db$Enrichment_cell_types)

imm.db %>% 
  ggplot(aes(y = Enrichment_cell_types, fill = Enrichment_cell_types)) +
  geom_bar(show.legend = F) +
  labs(x = "Number of datasets") +
  ggprism::theme_prism()
# there are 25 datasets for unbiased, we'll use those

sd <- imm.db %>% 
  filter(Enrichment_cell_types == "unbiased") %>%
  ggplot(aes(x = Cancer_type, y = N_samples_light,
             fill = Cancer_type)) +
  geom_col(aes(group = CSI_Dataset_id),
           position = position_dodge(width = .9),
          show.legend = F) +
  geom_hline(yintercept = 10, linetype = 2) +
  geom_label(aes(label = N_samples_light,
                group = CSI_Dataset_id),
                position = position_dodge(width = .9),
             fill = "white",
             show.legend = F) +
  labs(y = "Number of samples\n(small samples removed)") +
  ggprism::theme_prism() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# png("figs/samples_perCancer.png", width = 1500, height = 500)
 sd
# dev.off()
```

Now we will load the light (no small samples and downsampled) rds file for this corresponding files

```{r}
rds <- imm.db %>% 
        # get only whole tumor
        filter(Enrichment_cell_types == "unbiased") %>% 
        # get only datasets with >+ 10 samples/patients
        filter(N_samples_light >= 10) %>% 
        select(RDS_id, CSI_Dataset_id) %>% 
        as.data.frame()

```

## Load immucanscdb selected datasets
```{r}
# path to light rds
light.path <- "~/Dropbox/CSI/Standardized_SingleCell_Datasets/immucanscdb/output/light_rds"

#create a list to store the seurat objects
seu <- list()

for(a in 1:nrow(rds)){
  # look for the rds in the light rds directory
  rds.file <- list.files(light.path, rds[a,1], full.names = T)
  #load file
  tmp <- readRDS(rds.file)
  
  message("##### Loaded ", rds[a,2], "#####\n")
  
  # add missing metadata for CSI_Dataset_id
  tmp$CSI_Dataset_id <- rds[a,2]
  
  # add to list
  seu[[rds[a,2]]] <- tmp
}

rm(tmp)
gc()
```

## Load also Joanito dataset
```{r}
seu[["JoaI_2022_35773407"]] <- 
  readRDS("~/Dropbox/CSI/Standardized_SingleCell_Datasets/JoaI_2022_35773407/output/JoaI_2022_35773407_Nofilt_Downsampled_light.rds")

seu[["JoaI_2022_35773407"]]$CSI_Dataset_id <- "Joanito_2022_35773407"
```

## Explore list metadata
```{r}
wmeta.list <- list()
for(i in names(seu)){
  met <- seu[[i]]@meta.data
  
  names(met) <- tolower(names(met))
  
  if (!(any(grepl("anno", names(met))))){
    met <- met %>% mutate(annotation_authors = cell.type)
  } else if(!(any(grepl("annotation_authors", names(met))))){
    met <- met %>% mutate(annotation_authors = annotation_major)
  }
  
    if(any(grepl("sample$", names(met)))){
        sample.id <- "sample"
    } else if(any(grepl("sample.id", names(met)))) {
        sample.id <- grep("sample.id", names(met),
                           ignore.case = T, value = T)
        
    } else if (any(grepl("patient$", names(met)))) {
          sample.id <- "patient"

    } else if(any(grepl("patient", names(met)))) {
        sample.id <- grep("patient", names(met),
                           ignore.case = T, value = T)
    } else {
      sample.id <- "orig.ident"
    }
  
  cat(sample.id, "\n" )
  met$group <- as.character(met[,sample.id])
  
  wmeta.list[[i]] <-  met
}

rm(met)

wmeta <- wmeta.list %>% 
          data.table::rbindlist(use.names = T, fill = T)
rm(wmeta.list)
```

Plot

```{r}
sum.ct <- 
  wmeta %>% 
  # convert variant names
  mutate(annotation = StandardizeCellNames(annotation_authors)) %>% 
  count(group, annotation) %>% 
  group_by(group) %>% 
  mutate(Freq = n / sum(n)) 
  
```

```{r}
  
ctplot <- 
  sum.ct %>% 
  ggplot(aes(annotation, Freq, color = group)) +
    geom_boxplot(aes(group = annotation, color = group),
                show.legend = F, outlier.color = NA) +
    geom_point(show.legend = F, alpha = 0.5) +
    labs(y = "Relative proportion of authors' cell type\nannotation per sample",
         x = "Authors` annotation") +
    ggprism::theme_prism() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45,
                                     hjust =1, vjust = 1))

png("figs/authors_props_sample.png", height = 700, width = 1400)
ctplot
dev.off()
```


# Hitme annotation
play with S4 object



```{r}
scGate_models_DB <- scGate::get_scGateDB(branch = "master", force_update = TRUE)
models.TME <- scGate_models_DB$human$TME_HiRes

path_ref <- "~/Dropbox/CSI/reference_atlases" 
ref.maps <- list(CD8 = load.reference.map(file.path(path_ref, "CD8T_human_ref_v1.rds")),
          CD4 = load.reference.map(file.path(path_ref, "CD4T_human_ref_v2.rds")), 
           DC = load.reference.map(file.path(path_ref, "DC_human_ref_v1.rds")), 
          MoMac = load.reference.map(file.path(path_ref, "MoMac_human_v1.rds"))) 
```


```{r}

met2 <- annotate_cells(object = met,
                       scGate.model = models.TME,
                       ref.maps = ref.maps,
                       ncores = 9)


annotate_cells(object = met,
               dir = "haha")
```


```{r}
setClass(Class = "HiT",
         slots = list(
           metadata = "data.frame",
           Predictions = "list",
           Proportions = "list",
           Avg.Expression = "list",
           version = "list"
           )
         )

setClass(Class = "scGate",
         slots = list(
           UCell_score = "data.frame",
           scGate_is.pure = "data.frame",
           scGate_multi = "data.frame",
           additional.signatures = "data.frame"
           )
         )

setClass(Class = "map.ref",
         slots = list(
            Classifier = "data.frame",
            map.version = "character"
           )
         )

setClass(Class = "Projectils",
         slots = list(
           CD8 = "map.ref",
           CD4 = "map.ref",
           DC = "map.ref",
           MoMac = "map.ref"
           )
         )


```

```{r}
scgate <- new(Class = "scGate",
              UCell_score = met2@meta.data[,grep("_UCell$", names(met2@meta.data))],
              scGate_is.pure = met2@meta.data[,grep("^is.pure_", names(met2@meta.data))],
              scGate_multi = met2@meta.data[,grep("scGate_multi", names(met2@meta.data)), drop = F]
              )





hit <- new(Class = "HiT",
           metadata = met2@meta.data,
           Predictions = list("scGate" = scgate))
```












