---
title: Trial of HiTME package with whole tumor datasets
author: "J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---


# Prepare packages and environtment
```{r load_packages_functions, message=F, warning=F}
renv::restore()

# remotes::install_github("carmonalab/HiTME",
#                         ref="master",
#                         auth_token = "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")

# load packages
# pckgs <- c("tidyverse", "Seurat", "scGate",
#            "Matrix", "patchwork",
#            "data.table", "STACAS",
#            "SignatuR", "ProjecTILs",
#            "scCustomize", "UCell", "ProjecTILs")

pckgs <- c("tidyverse", "ProjecTILs", "HiTME"
           )
for(package in pckgs){
  tryCatch(
    {
      library(package,
              character.only = T,
              quietly = T)
    },
    error = function(e){
      message(e)
    })
}

options(timeout=4000) # increase time out for downloading

# source("R/utils.R")
# source("R/main.R")


#dataset testing
#seu <- list("Galen_2019_30827681" = readRDS("data/galen_Test.rds"))

my_seed <- 321
set.seed(my_seed)

```

Also, increase memory parameters to allocate large vectors, otherwise R might crash
```{bash create_Renviron}
#include vector max size
max_size=250Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Checking size of the allowed vector:
  cat .Renviron
fi
```




# Load data
We will use the datasets obtained from https://immucanscdb.vital-it.ch/ and previously processed. Only using the ones with whole tumor cells. Moreover, we will also add the Joanito et al., 2022 dataset which include whole tissue from colorectal samples.


The list of datasets from https://immucanscdb.vital-it.ch/ are detailed in https://docs.google.com/spreadsheets/d/1bFZ_jnmWUDJ9gaycsLaIsRmUryUWpdpW5aZY7V4Vpc0/edit#gid=0

We retrieve this dataframe from google drive

```{r load_db_specs}
imm.db <- readxl::read_xlsx("data/Standarized_datasets_DB.xlsx")
```


```{r, eval = F}
# keep whole tumor datasets
table(imm.db$Enrichment_cell_types)

imm.db %>% 
  ggplot(aes(y = Enrichment_cell_types, fill = Enrichment_cell_types)) +
  geom_bar(show.legend = F) +
  labs(x = "Number of datasets") +
  ggprism::theme_prism()
# there are 25 datasets for unbiased, we'll use those

sd <- imm.db %>% 
  filter(Enrichment_cell_types == "unbiased") %>%
  ggplot(aes(x = Cancer_type, y = N_samples_light,
             fill = Cancer_type)) +
  geom_col(aes(group = CSI_Dataset_id),
           position = position_dodge(width = .9),
          show.legend = F) +
  geom_hline(yintercept = 10, linetype = 2) +
  geom_label(aes(label = N_samples_light,
                group = CSI_Dataset_id),
                position = position_dodge(width = .9),
             fill = "white",
             show.legend = F) +
  labs(y = "Number of samples\n(small samples removed)") +
  ggprism::theme_prism() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# png("figs/samples_perCancer.png", width = 1500, height = 500)
 sd
# dev.off()
```

Now we will load the light (no small samples and downsampled) rds file for this corresponding files

```{r}
rds <- imm.db %>%
        # get only whole tumor
        filter(Enrichment_cell_types == "unbiased") %>% 
        # get only datasets with >+ 10 samples/patients
        filter(N_samples_light > 9) %>% 
        select(RDS_id, CSI_Dataset_id) %>% 
        as.data.frame()

```

## Load immucanscdb selected datasets
```{r}
# path to light rds
light.path <- "~/Dropbox/CSI/Standardized_SingleCell_Datasets/immucanscdb/output/light_rds"

#create a list to store the seurat objects
seu <- list()

for(a in 1:nrow(rds)){
  # look for the rds in the light rds directory
  rds.file <- list.files(light.path, rds[a,1], full.names = T)
  #load file
  tmp <- readRDS(rds.file)
  
  message("##### Loaded ", rds[a,2], "#####\n")
  
  # add missing metadata for CSI_Dataset_id
  tmp$CSI_Dataset_id <- rds[a,2]
  
  tmp[["RNA"]]@meta.features <- data.frame(row.names = rownames(tmp))
  
  # add to list
  seu[[rds[a,2]]] <- tmp
}

rm(tmp)
gc()
```

## Load also Joanito dataset
```{r, eval = F}
seu[["JoaI_2022_35773407"]] <- 
  readRDS("~/Dropbox/CSI/Standardized_SingleCell_Datasets/JoaI_2022_35773407/output/JoaI_2022_35773407_Nofilt_Downsampled_light.rds")

seu[["JoaI_2022_35773407"]]$CSI_Dataset_id <- "Joanito_2022_35773407"
```

## Explore list metadata
```{r}
wmeta.list <- list()
for(i in names(seu)){
  met <- seu[[i]]@meta.data
  
  names(met) <- tolower(names(met))
  
  if (!(any(grepl("anno", names(met))))){
    met <- met %>% mutate(annotation_authors = cell.type)
  } else if(!(any(grepl("annotation_authors", names(met))))){
    met <- met %>% mutate(annotation_authors = annotation_major)
  }
  
    if(any(grepl("sample$", names(met)))){
        sample.id <- "sample"
    } else if(any(grepl("sample.id", names(met)))) {
        sample.id <- grep("sample.id", names(met),
                           ignore.case = T, value = T)
        
    } else if (any(grepl("patient$", names(met)))) {
          sample.id <- "patient"

    } else if(any(grepl("patient", names(met)))) {
        sample.id <- grep("patient", names(met),
                           ignore.case = T, value = T)
    } else {
      sample.id <- "orig.ident"
    }
  
  cat(sample.id, "\n" )
  met$group <- as.character(met[,sample.id])
  
  wmeta.list[[i]] <-  met
  seu[[i]]@meta.data <- met %>% 
              mutate(annotation = HiTME:::StandardizeCellNames(annotation_authors))
  
}

rm(met)

wmeta <- wmeta.list %>% 
          data.table::rbindlist(use.names = T, fill = T)
rm(wmeta.list)
```

Plot

```{r, eval = F}
sum.ct <- 
  wmeta %>% 
  # convert variant names
  mutate(annotation = StandardizeCellNames(annotation_authors)) %>% 
  count(group, annotation) %>% 
  group_by(group) %>% 
  mutate(Freq = n / sum(n)) 
  
```

```{r, eval = F}
  
ctplot <- 
  sum.ct %>% 
  ggplot(aes(annotation, Freq, color = group)) +
    geom_boxplot(aes(group = annotation, color = group),
                show.legend = F, outlier.color = NA) +
    geom_point(show.legend = F, alpha = 0.5) +
    labs(y = "Relative proportion of authors' cell type\nannotation per sample",
         x = "Authors` annotation") +
    ggprism::theme_prism() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45,
                                     hjust =1, vjust = 1))

png("figs/authors_props_sample.png", height = 700, width = 1400)
ctplot
dev.off()
```


# Hitme annotation

## Load ref

```{r}

scGate_models_DB <- scGate::get_scGateDB(branch = "dev", force_update = F)
# models.TME <- scGate_models_DB$human$TME_HiRes

path_ref <- "~/Dropbox/CSI/reference_atlases" 
ref.maps <- list(CD8 = ProjecTILs::load.reference.map(file.path(path_ref, "CD8T_human_ref_v1.rds")),
          CD4 = ProjecTILs::load.reference.map(file.path(path_ref, "CD4T_human_ref_v2.rds")), 
           DC = ProjecTILs::load.reference.map(file.path(path_ref, "DC_human_ref_v1.rds")), 
          MoMac = ProjecTILs::load.reference.map(file.path(path_ref, "MoMac_human_v1.rds"))) 

# add scGate_link to ref.maps
layer1.links <- list("CD8" = "CL:0000625",
                  "CD4" = "CL:0000624",
                  "DC" = "CL:0000451",
                  "MoMac" = c("CL:0000576_CL:0000235")
                  )
for(a in names(ref.maps)){
  ref.maps[[a]]@misc$layer1_link <- layer1.links[[a]]
}
```

# Loop for all samples
```{r}
seu.list <- list()
hit.list <- list()

stime <- data.frame(matrix(nrow=0, ncol=9))
names(stime) <- c("Dataset", "Sample", "N_cells", "ncores", "user.self",
                  "sys.self","elapsed", "user.child", "sys.child" )
```


```{r}
ncores = 6

for(a in names(seu)){
  # split into samples/patients
  if(!all(unique(seu[[a]]$group) %in% names(hit.list))){
  dataset.list <- Seurat::SplitObject(seu[[a]], split.by = "group")
  
  for(b in names(dataset.list)){
    if(!b %in% names(seu.list)){
      
    # Run annotate cells
    tryCatch(
      {
      message("Running Run.HiTME dataset", a, ". Sample: ", b)
      
      time <- system.time(
        
      seu.list[[b]] <- Run.HiTME(dataset.list[[b]],
                                 scGate.model.branch = "dev",
                                 ref.maps = ref.maps,
                                return.Seurat = T,
                                ncores = ncores)
        
      )
      
      time <- time %>% as.data.frame() %>%
              t() %>% as.data.frame() %>%
              remove_rownames() %>% 
              mutate(Dataset = a,
                     Sample = b,
                     N_cells = ncol(seu.list[[b]]),
                     ncores = ncores)
      print(time)
      
      stime <- rbind(stime, time)
    
    # Get Hit object
    message("Creating Hit Object dataset", a, ". Sample: ", b)
    hit.list[[b]] <- get.HiTObject(seu.list[[b]])
      },
    error = function(e){
      message(e)
      message("Processing not possible for ", a, ". Dataset ", b)
      
    })
    } else {
      message(b, " already done! :). For dataset", a)
    }
  }
  } else {
    message(a, " dataset already done! :)")
  }
}

```


```{r}
saveRDS(seu.list, "cache/20datasets_annotated.rds")
saveRDS(hit.list, "cache/20hit_annotated.rds")
```



## Test new version
```{r}
require(HiTME)
sys.Yuan <- system.time(
test <- Run.HiTME(seu$Galen_2019_30827681[,1:5000],
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  ref.maps = ref.maps,
                  ncores = 9)
)

#.bpstop_impl(param)

hit <- get.HiTObject(test,
                     group.by = list("layer1" = "scGate_multi",
                                          "layer2" = "functional.cluster",
                                     "layer1.2" =  "annotation_chetah",
                                     "layer0" = "baba"
                                          )
                     )

avg <- get.aggregated.profile(test,
                                group.by.aggregated = c("scGate_multi",
                                                        "functional.cluster"),
                              gene.filter = c("IL10", "CXCR5"),
                              useNA = T)

prop <- get.celltype.composition(test,
                                 group.by.composition = list("eso" = "scGate_multi",
                                                          "peroque" = "functional.cluster",
                                                          "baba"),
                                 split.by = "group")

agg.sig <- get.aggregated.signature(test,
                                group.by.aggregated = c("scGate_multi", "functional.cluster", "baba"), fun = sum)
```

# Test v2
```{r}
add.sig <- GetSignature(SignatuR$Hs$Blocklists)
t0 <- Seurat::SplitObject(seu$Galen_2019_30827681, split.by = "group")

scGate_models_DB <- scGate::get_scGateDB(branch = "dev", force_update = F)
models.TME <- scGate_models_DB$human$TME_HiRes


# one object, for one sample

# scGate, projectils, return Seurat
test <- Run.HiTME(t0$aml210ad0,
                  scGate.model.branch = "dev",
                  return.Seurat = T,
                  ref.maps = ref.maps,
                  ncores = 9, additional.signatures = NULL)

aggr <- get.aggregated.profile(test,
                              group.by.aggregated = list("layer1" = "scGate_multi",
                                                          "layer2" = "functional.cluster",
                                                         "layer1.2" =  "annotation_chetah",
                                                         "layer0" = "baba"
                                                          ),
                              GO_accession = "GO:0007049")


test <- scGate::scGate(t0$aml1012d0,
               model =  scGate_models_DB$human$HiTME,
               ncores = 9)

# no scGate, projectils, return Seurat
test <- Run.HiTME(t0$aml1012d0,
                  scGate.model.branch = "dev",
                  return.Seurat = T,
                  ref.maps = NULL,
                  ncores = 9)

# scGate, no projectils, return Seurat
test <- Run.HiTME(t0$aml1012d0,
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  ref.maps = NULL,
                  additional.signatures = NULL,
                  ncores = 9)

# scGate, no projectils, return Seurat, no additional sigs
test <- Run.HiTME(t0$aml1012d0,
                  scGate.model.branch = "dev",
                  return.Seurat = T,
                  ref.maps = NULL,
                  additional.signatures = NULL,
                  ncores = 9)


View(test@meta.data)
```

```{r}
# list
test1 <- Run.HiTME(t0[4:6],
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  ref.maps = ref.maps,
                  ncores = 9)

test2 <- Run.HiTME(t0[4:6],
                  scGate.model = NULL,
                  return.Seurat = F,
                  ref.maps = NULL,
                  ncores = 9, useNA = T)

test.by <- Run.HiTME(t0[9:11],
                  scGate.model = NULL ,
                  return.Seurat = F,
                  ref.maps = list(ref.maps[[2]], ref.maps[[3]]),
                  ncores = 9,
                  useNA = T)

sp <- scCustomize::Merge_Seurat_List(t0[7:8])

object <- Run.HiTME(sp,
                    split.by = "group",
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  remerge = F,
                  ref.maps = ref.maps[2:3],
                  ncores = 9, additional.signatures = NULL)
hit.list <- list()
for(a in names(testby)){
 hit.list[[a]] <- get.HiTObject(testby[[a]],
                                group.by = list("layer1" = "scGate_multi",
                                          "layer2" = "functional.cluster",
                                         "layer1.2" =  "annotation_chetah",
                                         "layer0" = "baba"
                                          )) 
}

group.by = list("layer1" = "scGate_multi",
                                          "layer2" = "functional.cluster",
                                         "layer1.2" =  "annotation_chetah",
                                         "layer0" = "baba"
                                          )
com <- get.celltype.composition(object,
                                group.by.composition = group.by)

avg <- get.aggregated.profile(object,
                              group.by.aggregated = group.by)

sigs <- get.aggregated.signature(object,
                                 group.by.aggregated = group.by)

hit <- get.HiTObject(object,
              group.by = group.by)
```
# tests hit objects
```{r}
hit <- get.HiTObject(test,
                     group.by = list("layer1" = "scGate_multi",
                                          "layer2" = "functional.cluster",
                                     "layer1.2" =  "annotation_chetah",
                                     "layer0" = "baba"
                                          ))

hit <- Run.HiTME(t0$aml328d113,
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  ref.maps = ref.maps,
                  ncores = 9)

object <- Run.HiTME(t0$aml1012d0,
                  scGate.model.branch = "dev",
                  return.Seurat = T,
                  ref.maps = ref.maps,
                  ncores = 9)

avg <- get.aggregated.profile(object,
                                group.by.aggregated = c("scGate_multi",
                                                        "functional.cluster"),
                              gene.filter = c("IL10", "CXCR5"),
                              useNA = T)

head(avg$Average$scGate_multi$Ribosomal)

sigs <- get.aggregated.signature(object,
                                group.by.aggregated = c("scGate_multi",
                                                        "functional.cluster"),
                              useNA = F)
calc <- get.celltype.composition(test,
                                 group.by.composition = c("scGate_multi",
                                                        "functional.cluster"))
```

# test additional param
```{r}

hit.na <- Run.HiTME(t0$aml328d113,
                  scGate.model.branch = "dev",
                  return.Seurat = F,
                  ref.maps = ref.maps,
                  ncores = 9, useNA = T)
```


# Cell type consistency analysis
Load hit files
```{r load_db_specs}
imm.db <- readxl::read_xlsx("data/Standarized_datasets_DB.xlsx")
```

```{r}
require(tidyverse)
hit <- readRDS("cache/20hit_annotated.rds")

names(imm.db) <- tolower(names(imm.db))

# accommodate metdata
for(a in names(hit)){
  hit[[a]]@metadata <- left_join(hit[[a]]@metadata,
                                 imm.db, by = "csi_dataset_id")
}
```

```{r}
metadata <- c("cancer_type", "treatment_type")
t <- get.PCA.samples(hit, metadata = metadata)


matrix <- t$aggregated$layer1$`GO:0004896_Cytokine_activity`
metadata <- t$aggregated$layer1$metadata
silh <- cluster::silhouette(as.numeric(as.factor(metadata$celltype)),
                            stats::dist(t(matrix)))
print(factoextra::fviz_silhouette(silh)+
                            theme(legend.position = "none"))

mean(pl$data$sil_width)
```


```{r}
group.by <- c("sample", "celltype")
param <- BiocParallel::MulticoreParam(workers =  8, progressbar = T)
cc <- get.cluster.score(matrix,
                  metadata,
                  group.by = group.by, bparam = param)
```











